Investigating social change during cultural contact period using geometric morphometry of pottery shapes from Iron Age northeastern Taiwan

### 2020-01-26 analysis/paper/paper.Rmd
para 2 & 3

The emergence of social inequality in the context of cross-cultural interaction especially in a colonial situation was observed in many parts of world when the introduction of foreign trade goods to local indigenous societies. The monopolization of long-distance trade goods usually cause substantial transformations of indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Recently indirect effects of colonialism or pericolonial archaeology were address when investigating areas where European colonial rule was limited or their conquests were unsuccessful, but their colonial activities had economically and politically impacts on indigenous peoples in the periphery of colonial control [@Trabert2017; @Acabado2017]. For example, @Acabado2017's study on Ifugao society in the highland Philippines suggests economic and political intensification during the Spanish presence as the response of indigenous peoples to the Spanish cooptation. Indigenous societies might not just passively accept the colonial rules, but instead actively negotiated with the colonist, and accommodate or resist the foreign intrusion through their daily cultural practices, such as consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. Northeastern Taiwan is an ideal example to explore the indirect impacts of the foreign occupations on local indigenous societies due to its isolated location.  The archaeological evidence in northeastern Taiwan show evidence of increasing use of prestige goods such as trade ornaments and uneven distribution pattern in domestic area and burial contexts when the society encountered by the Spanish and the Dutch in the 17th century. In addition to the differential distribution of trade ornaments, the locally made ceramic present a high consistency in form and shape compared to other pottery throughout northeastern Taiwan that hints the possibility of pottery specialization.

To identify craft specialization, the variation of standardization of specialized product is commonly used as an indicator based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, increased skills, and fewer number of producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For ceramic standardization, several measurements have been proposed and discussed, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurement is widely applied to archaeological assemblages. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be assessed by comparing coefficient of variation and the differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts. The aim of this paper is to investigate if there is increasing ceramic specialization resulting from interaction with the Europeans in the 17th century or the Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies.

>>>>

The emergence of social inequality in the context of cross-cultural interaction, especially in a colonial situation, has been observed in many parts of world, for example, when foreign trade goods are introduced to local Indigenous societies. The monopolization of long-distance trade goods has caused substantial transformations of Indigenous economic, cultural, and socio-political systems [@Dietler2005; @Dietler1997; @Junker1993; @Silliman2005]. Pericolonial archaeology is the study of these indirect effects of colonialism, investigating areas where European colonial rule was limited and their conquests were short-lived and unsuccessful, but their colonial activities had economic and political impacts on Indigenous peoples living on the periphery of colonial control and influence [@Trabert2017; @Acabado2017]. @Acabado2017's study of Ifugao society in the  Philippines highland suggests economic and political intensification during the Spanish presence as the response of Indigenous peoples to the Spanish cooptation. Indigenous societies might not simply passively accept colonial rule, but instead actively negotiate with the colonist, and accommodate or resist foreign intrusion through their daily cultural practices, such as their consumption patterns of foreign goods [@Dietler2015; @Given2004; @Mullins2011; @Scaramelli2005; @Silliman2001; @TorrenceandClarke2000]. In this paper we present a case study from Northeastern Taiwan as an example of these indirect impacts of foreign occupations on local Indigenous societies. Our aim is to investigate if there was increasing ceramic specialization resulting from Indigenous interaction with Europeans in the 17th century or Chinese in the 19th century, two major foreign influences in early historical Taiwan, that might indicate social changes in local indigenous societies. Previous reports of archaeological evidence in northeastern Taiwan descrbes increases in the use of prestige goods, such as trade ornaments, and uneven artefact distribution patterns in domestic areas and burial contexts at the time when the Indigenous communities encountered the Spanish and the Dutch in the 17th century. Here we show that in addition to the differential distribution of trade ornaments post-European contact, the locally-made ceramics become highly consistent in their form and shape, compared to other pottery throughout northeastern Taiwan. This shift to higher consistency hints at the possibility of craft specialization in pottery manufacture.

To identify craft specialization in the archaeological record, variation in the standardization of a specialized product is often used as an indicator. This is based on the assumption that specialized mass production will lead to uniformity of the product due to routinization, higher skills, and fewer producers involved in production [@Stark1995; @Costin1991; @Costin2001]. For investigating ceramic standardization, several measurements have been discussed and used, such as metric, compositional, and technological approaches [@Arnold2000; @Boness2015; @Rice1991; @Blackman1993; @Costin1991; @Tite1999; @Roux2015]. Among those variables, metric measurements are most widely applied to archaeological assemblages because of the ease of collecting these data. By comparing variations for sets of data using coefficient of variation coupled with statistical testing, the degree of standardization can be identified according to values [@Eerkens2001; @Roux2003; @Stark1995]. @Roux2018's ethnographic study in the Jodhpur region of India suggests the number of artisans can be statistically assessed by comparing coefficients of variation and differences between artisans can be detected at both intra-individual level and inter-individual level in a region. However, linear measurements might be limited for identifying differences in variations because they could be insensitive to subtle variations resulting from changes in assemblages. Shape analysis based on geometric morphometrics method are new approaches for measuring the degree of standardization since it takes the overall shape as variable for comparison [@Slice2007]. Using geometric morphometrics approaches, we study shape and standardization of locally made ceramics at Kiwulan (1350-1950 AD) [@Chen2007], a large multi-component archaeological site in northeastern Taiwan, to identify changes in ceramic production in cross-cultural interaction contexts.
